Symbolic Weighted Language Models and\\ Parsing over Infinite Alphabets

### 2021-07-01 introduction.tex
1 passe intro legere

% !TEX root = main.tex
%\todo
% Introduction
%
Parsing %variant of membership problem for formal languages
%- given a language model and an input word $s$, compute a derivation of the model that yields $s$.
is the problem %process 
of structuring a linear representation in input %typically
(a finite word) according to a language model. % (a formal grammar).
% natural language, programming language, 
%
Most of the context-free parsing approaches~\cite{GruneJacobs08parsing}  
assume a finite and reasonably small input alphabet. %models and algorithms 
Such a restriction makes perfectly sense in the context of NLP tasks 
like constituency parsing or of programming languages compilers or interpreters.
Considering large or infinite alphabets can however be of practical interest. %in other cases.
%
For instance, when dealing with large characters encodings such as UTF-16, % processing strings in 
\eg for vulnerability detection in Web-applications~\cite{dAntoni21CACM}, 
%
for the analyse (\eg validation or filtering) 
of data streams or serialization of structured documents 
(which may contain textual or numerical attributes)~\cite{Segoufin06csl}, 
or for processing timed execution traces~\cite{Bouyer03algebraic}.
% algebraic definition of a class of data languages 
% (notion of monoid recognizability, based on registers, comparable to Bojancszik et al. data words)
The latter is related to a case study that motivated the present work: 
automated music transcription.
In this problem, a symbolic representation of a music performance, % a symbolic representation of a music performance 
in the form of a sequence of timed musical events, % timestamped
is converted into a structured score in Common Western Music Notation~\cite{Gould11Notation}.
It can therefore be stated as a parsing problem~\cite{foscarin:hal-01988990}, 
over an infinite alphabet of timed events.

Various extensions of language models for handling infinite alphabets have been studied.
%words carrying data values in an infinite domain (e.g. integers) e.g. data processing 
For instance, some automata with memory extensions 
allow restricted storage and comparison of input symbols, 
%and correspond with logics, 
(see~\cite{Segoufin06csl} for a survey),
\mynote{register: skip refs and details, add Mikolaj}
with pebbles for marking positions~\cite{NevenSchwentickVianu04FSMinfinite}, 
registers~\cite{KaminskiFrancez94}, 
or %computations in 2 steps
the possibility to compute on subsequences 
with the same attribute values~\cite{Bojanczyk11FO2}. % data words automata.
%
%for the and verification of infinite state systems 
%(model checkers: alphabet = long bit-vectors)
%...For the representation of  in model checking, verification and 
The automata at the core of model checkers
compute on input symbols represented by large bitvectors~\cite{Vardi07ciaa} %\cite{BaierKatoen08MC}
(sets of assignments of Boolean variables) %propositional variables)
and in practice,  %implementation
each transition accepts a set of such symbols (instead of an individual symbol), 
represented by Boolean formula or Binary Decision Diagrams.
%
Following a similar idea, % and generalizing,
in symbolic automata (\SA)~\cite{dAntoniVeanes17CAV,dAntoni21CACM}, 
the transitions are guarded by predicates over infinite alphabet domains.
With closure conditions on the sets of such predicates, % (alphabet theories), 
all the good properties enjoyed by automata over finite alphabets are preserved.

Other extensions of language models  %(automata and grammars) 
help in dealing with non-determinism, with the computation of weight values. 
With an ambiguous grammar, there may exists several derivations 
(\emph{abstract syntax trees} -- AST), % \ie the result of an analyze.
yielding one input word. % structuring a word, 
The association of one weight value %in semiring domain
to each AST permits to select a best one (or $n$ bests). % a fixed number of bests 
% = ...a ranking of ASTs 
This is roughly the principle of \emph{weighted parsing}
approaches~\cite{Goodman99SemiringParsing,Nederhof03weightedParsing,MorbitzVogler19weighted-parsing}.
In \emph{weighted language models}, 
like \eg probabilistic context-free grammars % (CFG),
and weighted automata (\WA) \cite{Droste09handbook},
a weight value is associated to each transition rule, % production rule
and the rule's weights can be combined with a associative product operator~$\otimes$ into 
the weight of a AST.
A second operator~$\oplus$, associative and commutative, 
is moreover used to handle ambiguity of the model, 
by summing the weights of the possibly several (in general exponentially many) AST %syntax trees 
associated to a given input word.
Typically, $\oplus$ will select the best of two weight values.
%Intuitively,~$\oplus$ selects, or ranking, the syntax trees.
The weight domain, equipped with these two operators shall be, at minima, 
a \emph{semiring} %$\Semiring$ 
where $\oplus$ can be extended to infinite sums, 
%\cite{Eilenberg74automata}
such as the Viterbi semiring and the tropical min-plus algebra, see Figure~\ref{fig:semirings}. 
%of domain $\mathbb{R}^+ \cup \{ +\infty\}$, 
%where $\oplus$ is min and $\otimes$ is plus .
%by ranking 
%making the weight domain a semiring.
%Some efficient specialized parsing algorithms~\cite{Huang05kbest} have been proposed in this context 
%% models represented as hypergraphs \cite{Huang05kbest}
%in order to compute the $n$ best syntax trees of a given input word without having to enumerate them all.
%Generally based on dynamic programming, these algorithms rely on 
%additional algebraic properties of~$\Semiring$.
%-- see \eg~\cite{Huang05kbest} for some NLP applications.
%The extraction of $n$ best list is useful 
%the problem: quantitative parsing or symbolic parsing
%of parsing of words over infinite input alphabet.

@@

It is based on \emph{symbolic weighted} finite states language models (\swM), generalizing 
both~\SA, with functions into an arbitrary semiring instead of Boolean guards, 

@@

In the transition rules of \swM models, input symbols appear as variables %(parameters)

@@

transducers (\SWT) and pushdown automata,

@@

a \SWT is used to define a distance $T(s, t)$ between two finite words $s$ and $t$

@@

a best search algorithm similar to shortest distance in graphs~\cite{Mohri02semiring,Huang05kbest}.

@@

% First one general edit-distance is defined by a weighted word 
% transducer~\cite{Mohri}
% %Symbolic automata transducers are extended models~\cite{VeanesdAntoniJACM}
% %dealing with infinite set of input symbols...
% value in a semiring...

>>>>

% !TEX root = main.tex
%\todo
% Introduction
%
Parsing %variant of membership problem for formal languages
%- given a language model and an input word $s$, compute a derivation of the model that yields $s$.
is the problem %process 
of structuring a linear representation on input %typically
(a finite word), according to a language model. % (a formal grammar).
% natural language, programming language, 
%
Most of the context-free parsing approaches~\cite{GruneJacobs08parsing}  
assume a finite and reasonably small input alphabet. %models and algorithms 
Such a restriction makes perfect sense in the context of NLP tasks such as constituency parsing,
or of programming languages compilers or interpreters.
Considering large or infinite alphabets can however be of practical interest, %in other cases.
%
for instance, when dealing with large characters encodings such as UTF-16, % processing strings in 
\eg for vulnerability detection in Web-applications~\cite{dAntoni21CACM}, 
%
for the analyse (\eg validation or filtering) 
of data streams or serialization of structured documents 
(with textual or numerical attributes)~\cite{Segoufin06csl}, 
or for processing timed execution traces~\cite{Bouyer03algebraic}.
% algebraic definition of a class of data languages 
% (notion of monoid recognizability, based on registers, comparable to Bojancszik et al. data words)
%
The latter case is related to a study that motivated the present work: 
automated music transcription.
In this problem, a music performance, % a symbolic representation of a music performance 
represented symbolically in the form of a sequence of timed musical events, % timestamped
is converted into a score in Common Western Music Notation~\cite{Gould11Notation},
structured according to nested grouping and metric strength of events.
%(the structure represents hierarchical  grouping of events, and their relative temporal positions)
It can therefore be stated as a parsing problem~\cite{foscarin:hal-01988990}, 
over an infinite alphabet of timed events.

Various extensions of language models for handling infinite alphabets have been studied.
%words carrying data values in an infinite domain (e.g. integers) e.g. data processing 
For instance, some automata with memory extensions 
allow restricted storage and comparison of input symbols, 
%and correspond with logics, 
(see~\cite{Segoufin06csl} for a survey),
\mynote{register: skip refs and details, add Mikolaj recent}
with pebbles for marking positions~\cite{NevenSchwentickVianu04FSMinfinite}, 
registers~\cite{KaminskiFrancez94}, 
or %computations in 2 steps
the possibility to compute on subsequences 
with the same attribute values~\cite{Bojanczyk11FO2}. % data words automata.
%
%for the and verification of infinite state systems 
%(model checkers: alphabet = long bit-vectors)
%...For the representation of  in model checking, verification and 
%
The automata at the core of model checkers
compute on input symbols represented by large bitvectors~\cite{Vardi07ciaa} %\cite{BaierKatoen08MC}
(sets of assignments of Boolean variables) %propositional variables)
and in practice,  %implementation
each transition accepts a set of such symbols (instead of an individual symbol), 
represented by Boolean formula or Binary Decision Diagrams.
%
Following a similar idea, % and generalizing,
in symbolic automata (\SA)~\cite{dAntoniVeanes17CAV,dAntoni21CACM}, 
the transitions are guarded by predicates over infinite alphabet domains.
With appropriate closure conditions on the sets of such predicates, % (alphabet theories), 
all the good properties enjoyed by automata over finite alphabets are preserved.

Other extensions of language models  %(automata and grammars) 
help in dealing with non-determinism, by the computation of weight values. 
With an ambiguous grammar, there may exist several derivations 
(\emph{abstract syntax trees} -- AST) % \ie the result of an analyze.
yielding one input word. % structuring a word, 
The association of one weight value %in semiring domain
to each AST permits to select a best one (or $n$ bests). % a fixed number of bests 
% = ...a ranking of ASTs 
This is roughly the principle of \emph{weighted parsing}
approaches~\cite{Goodman99SemiringParsing,Nederhof03weightedParsing,MorbitzVogler19weighted-parsing}.
In \emph{weighted language models}, 
like \eg probabilistic context-free grammars % (CFG),
and weighted automata (\WA) \cite{Droste09handbook},
a weight value is associated to each transition rule, % production rule
and the rule's weights can be combined with a associative product operator~$\otimes$ into 
the weight of an AST.
A second operator~$\oplus$, associative and commutative, 
is moreover used to handle the ambiguity of the model, 
by summing the weights of the possibly several (in general exponentially many) AST %syntax trees 
associated to a given input word.
Typically, $\oplus$ will select the best of two weight values.
%Intuitively,~$\oplus$ selects, or ranking, the syntax trees.
The weight domain, equipped with these two operators shall be, at minima, 
a \emph{semiring} %$\Semiring$ 
where $\oplus$ can be extended to infinite sums, 
%\cite{Eilenberg74automata}
such as the Viterbi semiring and the tropical min-plus algebra, see Figure~\ref{fig:semirings}. 
%of domain $\mathbb{R}^+ \cup \{ +\infty\}$, 
%where $\oplus$ is min and $\otimes$ is plus .
%by ranking 
%making the weight domain a semiring.
%Some efficient specialized parsing algorithms~\cite{Huang05kbest} have been proposed in this context 
%% models represented as hypergraphs \cite{Huang05kbest}
%in order to compute the $n$ best syntax trees of a given input word without having to enumerate them all.
%Generally based on dynamic programming, these algorithms rely on 
%additional algebraic properties of~$\Semiring$.
%-- see \eg~\cite{Huang05kbest} for some NLP applications.
%The extraction of $n$ best list is useful 
%the problem: quantitative parsing or symbolic parsing
%of parsing of words over infinite input alphabet.

@@

It is based on \emph{symbolic weighted} finite states language models (\swM), 
generalizing~\SA, with functions into an arbitrary semiring instead of Boolean guards, 

@@

In the transition rules of \swM models, input symbols appear as variables, %(parameters)

@@

transducers (\SWT), and pushdown automata

@@

a \SWT can define a distance $T(s, t)$ between finite words $s$ and $t$

@@

a best search algorithm similar to the shortest distance 
in graphs~\cite{Mohri02semiring,Huang05kbest}.

@@

% First one general edit-distance is defined by a weighted word 
% transducer~\cite{Mohri}
% %Symbolic automata transducers are extended models~\cite{VeanesdAntoniJACM}
% %dealing with infinite set of input symbols...
% value in a semiring...
\mynote{expressiveness: VPA have restricted equality test. 
        comparable to pebble automata? $\to$ conclusion}
